\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-07-22 analysis/paper/paper.tex

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}

@@

\usepackage{amsmath}

@@

\section{Code formatting}\label{code-formatting}

@@

\begin{itemize}
\tightlist
\item
  \proglang{Java}
\item
  \pkg{plyr}
\item
  \code{print("abc")}
\end{itemize}

\section{Introduction}\label{introduction}

@@

\section{Theoretical details}\label{theoretical-details}

@@

\section{Implementation details}\label{implementation-details}

@@

\subsection{Population-time plots}\label{population-time-plots}

\subsection{Data analysis}\label{data-analysis}

@@

\section{Case study 1--European Randomized Study of Prostate Cancer
Screening}\label{case-study-1european-randomized-study-of-prostate-cancer-screening}

@@

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
R+                        levels = c(0,1), 
R+                        labels = c("Control group", "Screening group"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> plot(pt_object)
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-2-1} \end{center}

@@

\begin{CodeInput}
R> pt_object_strat <- casebase::popTime(ERSPC, 
R+                                      event = "DeadOfPrCa", 
R+                                      exposure = "ScrArm")
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> plot(pt_object_strat)
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-3-1} \end{center}

@@

\begin{CodeInput}
R> plot(pt_object_strat, ncol = 2)
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-4-1} \end{center}

@@

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
R+                              data = ERSPC)
R> summary(cox_model)
\end{CodeInput}

\begin{CodeOutput}
Call:
survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
    ScrArm, data = ERSPC)

  n= 159893, number of events= 540

                        coef exp(coef) se(coef)     z Pr(>|z|)  
ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

                      exp(coef) exp(-coef) lower .95 upper .95
ScrArmScreening group     0.801       1.25     0.674     0.952

Concordance= 0.519  (se = 0.011 )
Rsquare= 0   (max possible= 0.075 )
Likelihood ratio test= 6.45  on 1 df,   p=0.0111
Wald test            = 6.37  on 1 df,   p=0.0116
Score (logrank) test = 6.39  on 1 df,   p=0.0115
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R+                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]))
R> legend("topleft", 
R+        legend = c("Control group", "Screening group"), 
R+        col = c("red","blue"),
R+        lty = c(1, 1), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-6-1} \end{center}

@@

\begin{CodeInput}
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
R+                                                   data = ERSPC, 
R+                                                   ratio = 100)
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> summary(casebase_exponential)
\end{CodeInput}

Call:
glm(formula = formula, family = binomial, data = sampleData)

Deviance Residuals: 
   Min      1Q  Median      3Q     Max  
-0.458  -0.458  -0.409  -0.409   2.247

Coefficients:
                      Estimate Std. Error z value            Pr(>|z|)    
(Intercept)            -7.7546     0.0584  -132.8 <0.0000000000000002 ***
ScrArmScreening group  -0.2397     0.0921    -2.6              0.0093 ** 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 3619.1  on 5939  degrees of freedom
Residual deviance: 3612.2  on 5938  degrees of freedom
AIC: 3616

Number of Fisher Scoring iterations: 5
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_exponential)[2])
\end{CodeInput}

\begin{CodeOutput}
ScrArmScreening group 
                 0.79 
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_exponential)[2,])
\end{CodeInput}

\begin{CodeOutput}
Waiting for profiling to be done...
\end{CodeOutput}

\begin{CodeOutput}
 2.5 % 97.5 % 
  0.66   0.94 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-8-1} \end{center}

@@

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
R+                                  data = ERSPC, 
R+                                  ratio = 100)
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> summary(casebase_time)
\end{CodeInput}

Call:
glm(formula = formula, family = binomial, data = sampleData)

Deviance Residuals: 
   Min      1Q  Median      3Q     Max  
-1.030  -0.489  -0.367  -0.278   2.740

Coefficients:
                      Estimate Std. Error z value            Pr(>|z|)    
(Intercept)            -9.0228     0.1186  -76.07 <0.0000000000000002 ***
Follow.Up.Time          0.2217     0.0159   13.97 <0.0000000000000002 ***
ScrArmScreening group  -0.2680     0.0939   -2.85              0.0043 ** 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 3619.1  on 5939  degrees of freedom
Residual deviance: 3402.7  on 5937  degrees of freedom
AIC: 3409

Number of Fisher Scoring iterations: 5
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
          (Intercept)        Follow.Up.Time ScrArmScreening group 
              0.00012               1.24824               0.76490 
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
Waiting for profiling to be done...
\end{CodeOutput}

\begin{CodeOutput}
                         2.5 %  97.5 %
(Intercept)           0.000095 0.00015
Follow.Up.Time        1.210257 1.28799
ScrArmScreening group 0.635606 0.91877
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-10-1} \end{center}

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
R+                                     data = ERSPC, 
R+                                     ratio = 100)
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> summary(casebase_splines)
\end{CodeInput}

Call:
glm(formula = formula, family = binomial, data = sampleData)

Deviance Residuals: 
   Min      1Q  Median      3Q     Max  
-1.245  -0.521  -0.397  -0.211   3.208

Coefficients:
                             Estimate Std. Error z value
(Intercept)                  -10.5377     0.3384  -31.14
splines::bs(Follow.Up.Time)1   4.9852     0.8551    5.83
splines::bs(Follow.Up.Time)2   1.7658     0.4963    3.56
splines::bs(Follow.Up.Time)3   5.1969     0.7203    7.22
ScrArmScreening group         -0.1850     0.0941   -1.97
                                         Pr(>|z|)    
(Intercept)                  < 0.0000000000000002 ***
splines::bs(Follow.Up.Time)1     0.00000000553602 ***
splines::bs(Follow.Up.Time)2              0.00037 ***
splines::bs(Follow.Up.Time)3     0.00000000000054 ***
ScrArmScreening group                     0.04937 *  
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 3619.1  on 5939  degrees of freedom
Residual deviance: 3352.4  on 5935  degrees of freedom
AIC: 3362

Number of Fisher Scoring iterations: 6
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
                 (Intercept) splines::bs(Follow.Up.Time)1 
                    0.000027                   146.229029 
splines::bs(Follow.Up.Time)2 splines::bs(Follow.Up.Time)3 
                    5.846329                   180.718284 
       ScrArmScreening group 
                    0.831083 
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
Waiting for profiling to be done...
\end{CodeOutput}

\begin{CodeOutput}
                                 2.5 %    97.5 %
(Intercept)                   0.000013   0.00005
splines::bs(Follow.Up.Time)1 28.719353 825.32637
splines::bs(Follow.Up.Time)2  2.210676  15.57574
splines::bs(Follow.Up.Time)3 44.370102 759.24168
ScrArmScreening group         0.690354   0.99867
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R+                                     time = seq(0,15,0.1), 
R+                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-12-1} \end{center}

@@

\begin{CodeInput}
R> anova(casebase_time, casebase_splines, test = "LRT")
\end{CodeInput}

\begin{CodeOutput}
Analysis of Deviance Table

Model 1: DeadOfPrCa ~ Follow.Up.Time + ScrArm + offset(offset)
Model 2: DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm + offset(offset)
  Resid. Df Resid. Dev Df Deviance       Pr(>Chi)    
1      5937       3403                               
2      5935       3352  2     50.3 0.000000000012 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

@@

\section{Case study 2--Bone-marrow
transplant}\label{case-study-2bone-marrow-transplant}

@@

\begin{CodeInput}
R> library(casebase)
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{casebase_jss_files/figure-latex/compPop1-1}

}

\caption{\label{fig:compPop1}Population-time plot for the stem-cell transplant study. The points represent the event of interest (i.e., relapse).}\label{fig:compPop1}
\end{figure}
\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{casebase_jss_files/figure-latex/compPop2-1}

}

\caption{\label{fig:compPop2}Population-time plot for the stem-cell transplant study. The points represent the competing events.}\label{fig:compPop2}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeInput}
R> model_cb <- fitSmoothHazard(
R+     Status ~ ftime + Sex + D + Phase + Source + Age, 
R+     data = bmtcrr, 
R+     ratio = 100, 
R+     time = "ftime")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{tabular}{l|r|l}
\hline
Covariates & HR & 95\% CI\\
\hline
Sex & 0.72 & (0.4, 1.29)\\
\hline
Disease & 0.61 & (0.33, 1.15)\\
\hline
Phase (CR2 vs. CR1) & 1.31 & (0.52, 3.31)\\
\hline
Phase (CR3 vs. CR1) & 1.58 & (0.39, 6.32)\\
\hline
Phase (Relapse vs. CR1) & 4.50 & (2.07, 9.77)\\
\hline
Source & 1.42 & (0.46, 4.41)\\
\hline
Age & 1.00 & (0.97, 1.02)\\
\hline
\end{tabular}

@@

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 100)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
R+                                      levels = levels(bmtcrr[,"Sex"])),
R+                       "D" = c("ALL", "AML"),
R+                       "Phase" = factor(c("Relapse", "Relapse"), 
R+                                        levels = levels(bmtcrr[,"Phase"])),
R+                       "Age" = c(35, 35),
R+                       "Source" = factor(c("PB", "PB"), 
R+                                         levels = levels(bmtcrr[,"Source"])))
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
R+                         method = "montecarlo", newdata = newdata)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-19-1}

}

\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-19}
\end{figure}
\end{CodeChunk}

\section{Case study 3}\label{case-study-3}

@@

\section{Discussion}\label{discussion}

@@

\bibliography{casebase_references.bib}

>>>>

\documentclass[
]{jss}

\usepackage[utf8]{inputenc}

@@

\usepackage{amsmath} \usepackage{longtable}

@@

\hypertarget{code-formatting}{%
\section{Code formatting}\label{code-formatting}}

@@

\begin{itemize}
\item
  \proglang{Java}
\item
  \pkg{plyr}
\item
  \code{print("abc")}
\end{itemize}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\begin{table}
  \begin{tabular}{p{1.8in}lll}
\hline
    &  Parameters &  Density \proglang{R} function & \code{dist}\\
    & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
\hline
    Exponential & \code{\color{red}{rate}}             & \code{dexp}   & \code{"exp"} \\
    Weibull (accelerated failure time)     & \code{shape, {\color{red}{scale}}}     & \code{dweibull} & \code{"weibull"} \\
    Weibull (proportional hazards)     & \code{shape, {\color{red}{scale}}}     & \code{dweibullPH} & \code{"weibullPH"} \\
    Gamma       & \code{shape, \color{red}{rate}}      & \code{dgamma} & \code{"gamma"}\\
    Log-normal  & \code{{\color{red}{meanlog}}, sdlog}   & \code{dlnorm} & \code{"lnorm"}\\
    Gompertz    & \code{shape, {\color{red}{rate}}}      & \code{dgompertz} & \code{"gompertz"} \\
    Log-logistic & \code{shape, {\color{red}{scale}}}   & \code{dllogis} & \code{"llogis"}\\
    Generalized gamma (Prentice 1975)   & \code{{\color{red}{mu}}, sigma, Q} & \code{dgengamma} & \code{"gengamma"} \\
    Generalized gamma (Stacy 1962)& \code{shape, {\color{red}{scale}}, k} & \code{dgengamma.orig} & \code{"gengamma.orig"} \\
    Generalized F     (stable)    & \code{{\color{red}{mu}}, sigma, Q, P} & \code{dgenf} & \code{"genf"} \\
    Generalized F     (original)  & \code{{\color{red}{mu}}, sigma, s1, s2} & \code{dgenf.orig} & \code{"genf.orig"} \\
\hline
  \end{tabular}
  \caption{Built-in parametric survival distributions in \pkg{flexsurv}.}
  \label{tab:dists}
\end{table}

\hypertarget{theoretical-details}{%
\section{Theoretical details}\label{theoretical-details}}

@@

\hypertarget{implementation-details}{%
\section{Implementation details}\label{implementation-details}}

@@

\hypertarget{population-time-plots}{%
\subsection{Population-time plots}\label{population-time-plots}}

\hypertarget{data-analysis}{%
\subsection{Data analysis}\label{data-analysis}}

@@

\hypertarget{case-study-1european-randomized-study-of-prostate-cancer-screening}{%
\section{Case study 1--European Randomized Study of Prostate Cancer
Screening}\label{case-study-1european-randomized-study-of-prostate-cancer-screening}}

@@

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
R>                        levels = c(0,1), 
R>                        labels = c("Control group", "Screening group"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
R> plot(pt_object)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> pt_object_strat <- casebase::popTime(ERSPC, 
R>                                      event = "DeadOfPrCa", 
R>                                      exposure = "ScrArm")
R> plot(pt_object_strat)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> plot(pt_object_strat, ncol = 2)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
R>                              data = ERSPC)
R> summary(cox_model)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R>                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]))
R> legend("topleft", 
R>        legend = c("Control group", "Screening group"), 
R>        col = c("red","blue"),
R>        lty = c(1, 1), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
R>                                                   data = ERSPC, 
R>                                                   ratio = 100)
R> 
R> summary(casebase_exponential)
R> exp(coef(casebase_exponential)[2])
R> exp(confint(casebase_exponential)[2,])
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R>                                           time = seq(0,15,0.1), 
R>                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
R>                                  data = ERSPC, 
R>                                  ratio = 100)
R> 
R> summary(casebase_time)
R> exp(coef(casebase_time))
R> exp(confint(casebase_time))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R>                                           time = seq(0,15,0.1), 
R>                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
R>                                     data = ERSPC, 
R>                                     ratio = 100)
R> 
R> summary(casebase_splines)
R> exp(coef(casebase_splines))
R> exp(confint(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R>                                     time = seq(0,15,0.1), 
R>                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> anova(casebase_time, casebase_splines, test = "LRT")
\end{CodeInput}
\end{CodeChunk}

@@

\hypertarget{case-study-2bone-marrow-transplant}{%
\section{Case study 2--Bone-marrow
transplant}\label{case-study-2bone-marrow-transplant}}

@@

\begin{CodeInput}
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

@@

@@

\begin{CodeInput}
R> model_cb <- fitSmoothHazard(
R>     Status ~ ftime + Sex + D + Phase + Source + Age, 
R>     data = bmtcrr, 
R>     ratio = 100, 
R>     time = "ftime")
\end{CodeInput}
\end{CodeChunk}

@@

@@

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 50)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
R>                                      levels = levels(bmtcrr[,"Sex"])),
R>                       "D" = c("ALL", "AML"),
R>                       "Phase" = factor(c("Relapse", "Relapse"), 
R>                                        levels = levels(bmtcrr[,"Phase"])),
R>                       "Age" = c(35, 35),
R>                       "Source" = factor(c("PB", "PB"), 
R>                                         levels = levels(bmtcrr[,"Source"])))
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
R>                         method = "numerical", newdata = newdata)
\end{CodeInput}
\end{CodeChunk}

\hypertarget{case-study-3}{%
\section{Case study 3}\label{case-study-3}}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@

\hypertarget{environment-details}{%
\section{Environment Details}\label{environment-details}}

This report was generated on 2019-07-22 17:17:57 using the following
computational environment and dependencies:

\begin{CodeInput}
R> # which R packages and versions?
R> devtools::session_info()
\end{CodeInput}
#> - Session info ----------------------------------------------------------
#>  setting  value                       
#>  version  R version 3.6.0 (2019-04-26)
#>  os       Pop!_OS 18.10               
#>  system   x86_64, linux-gnu           
#>  ui       X11                         
#>  language en_CA:en                    
#>  collate  en_CA.UTF-8                 
#>  ctype    en_CA.UTF-8                 
#>  tz       America/Toronto             
#>  date     2019-07-22                  
#> 
#> - Packages --------------------------------------------------------------
#>  package     * version    date       lib
#>  assertthat    0.2.1      2019-03-21 [1]
#>  backports     1.1.4      2019-04-10 [1]
#>  broom         0.5.1      2018-12-05 [1]
#>  callr         3.2.0      2019-03-15 [1]
#>  casebase    * 0.1.1.9000 2019-07-22 [1]
#>  cellranger    1.1.0      2016-07-27 [1]
#>  cli           1.1.0      2019-03-19 [1]
#>  colorspace    1.4-1      2019-03-18 [1]
#>  crayon        1.3.4      2017-09-16 [1]
#>  data.table    1.12.2     2019-04-07 [1]
#>  desc          1.2.0      2018-05-01 [1]
#>  devtools      2.0.1      2018-10-26 [1]
#>  digest        0.6.20     2019-07-04 [1]
#>  dplyr       * 0.8.0.1    2019-02-15 [1]
#>  evaluate      0.14       2019-05-28 [1]
#>  forcats     * 0.4.0      2019-02-17 [1]
#>  fs            1.3.1      2019-05-06 [1]
#>  generics      0.0.2      2018-11-29 [1]
#>  ggplot2     * 3.2.0      2019-06-16 [1]
#>  glue          1.3.1      2019-03-12 [1]
#>  gtable        0.3.0      2019-03-25 [1]
#>  haven         2.1.0      2019-02-19 [1]
#>  hms           0.4.2      2018-03-10 [1]
#>  htmltools     0.3.6      2017-04-28 [1]
#>  httr          1.4.0      2018-12-11 [1]
#>  jsonlite      1.6        2018-12-07 [1]
#>  knitr         1.23       2019-05-18 [1]
#>  lattice       0.20-38    2018-11-04 [5]
#>  lazyeval      0.2.2      2019-03-15 [1]
#>  lubridate     1.7.4      2018-04-11 [1]
#>  magrittr    * 1.5        2014-11-22 [1]
#>  Matrix        1.2-17     2019-03-22 [5]
#>  memoise       1.1.0      2017-04-21 [1]
#>  mgcv          1.8-28     2019-03-21 [5]
#>  modelr        0.1.4      2019-02-18 [1]
#>  munsell       0.5.0      2018-06-12 [1]
#>  nlme          3.1-140    2019-05-12 [5]
#>  pillar        1.4.2      2019-06-29 [1]
#>  pkgbuild      1.0.3      2019-03-20 [1]
#>  pkgconfig     2.0.2      2018-08-16 [1]
#>  pkgload       1.0.2      2018-10-29 [1]
#>  prettyunits   1.0.2      2015-07-13 [1]
#>  processx      3.3.1      2019-05-08 [1]
#>  ps            1.3.0      2018-12-21 [1]
#>  purrr       * 0.3.2      2019-03-15 [1]
#>  R6            2.4.0      2019-02-14 [1]
#>  Rcpp          1.0.1      2019-03-17 [1]
#>  readr       * 1.3.1      2018-12-21 [1]
#>  readxl        1.3.0      2019-02-15 [1]
#>  remotes       2.0.2      2018-10-30 [1]
#>  rlang         0.4.0      2019-06-25 [1]
#>  rmarkdown     1.14       2019-07-12 [1]
#>  rprojroot     1.3-2      2018-01-03 [1]
#>  rstudioapi    0.10       2019-03-19 [1]
#>  rticles       0.9.1      2019-07-22 [1]
#>  rvest         0.3.2      2016-06-17 [1]
#>  scales        1.0.0      2018-08-09 [1]
#>  sessioninfo   1.1.1      2018-11-05 [1]
#>  stringi       1.4.3      2019-03-12 [1]
#>  stringr     * 1.4.0      2019-02-10 [1]
#>  survival    * 2.43-3     2018-11-26 [5]
#>  testthat      2.1.1      2019-04-23 [1]
#>  tibble      * 2.1.3      2019-06-06 [1]
#>  tidyr       * 0.8.2      2018-10-28 [1]
#>  tidyselect    0.2.5      2018-10-11 [1]
#>  tidyverse   * 1.2.1      2017-11-14 [1]
#>  usethis       1.5.0.9000 2019-06-24 [1]
#>  VGAM          1.1-1      2019-02-18 [1]
#>  withr         2.1.2      2018-03-15 [1]
#>  xfun          0.8        2019-06-25 [1]
